---
title: "CDC Chronic Disease Indicators - CVD Prevalence Project"
author: "Lenio, Novick, Tanveer, Thompson"
date: "12 April 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
```

## 1. Load & Cache

The raw dataset is large (~300k rows). We load it once and cache it so
subsequent knits skip the file read entirely.

```{r load, cache=TRUE, cache.path="cache/"}
raw <- read.csv("U.S._Chronic_Disease_Indicators.csv", stringsAsFactors = FALSE)
cat("Loaded:", nrow(raw), "rows,", ncol(raw), "columns\n")
```

## 2. Preprocessing

### 2a. Drop Sparse & Unnecessary Columns

Keep only the columns needed for modeling. The secondary stratification fields
(`Stratification2/3`, `ResponseID`, ID columns, etc.) are empty or near-empty
across the full dataset and are dropped here.

```{r drop-columns, cache=TRUE, cache.path="cache/", dependson="load"}
keep_cols <- c(
  "YearStart", "YearEnd", "LocationAbbr",
  "Topic", "Question", "QuestionID",
  "DataValueType", "DataValue",
  "LowConfidenceLimit", "HighConfidenceLimit",
  "StratificationCategory1", "Stratification1"
)
clean <- raw[, keep_cols]
cat("Columns kept:", ncol(clean), "\n")
cat("Rows:", nrow(clean), "\n")
```

### 2b. Remove Suppressed Cells

Some cells have no `DataValue` because the sample was too small to report.
These are removed rather than imputed, as they represent structural missingness.

```{r filter-suppress, cache=TRUE, cache.path="cache/", dependson="drop-columns"}
clean <- clean[!is.na(clean$DataValue), ]
cat("Rows after removing suppressed values:", nrow(clean), "\n")
cat("Remaining NAs in DataValue:", sum(is.na(clean$DataValue)), "\n")
```

### 2c. Filter to Relevant Data Value Types

Feature indicators (tobacco, diabetes, obesity, CVD conditions) are reported as
Crude Prevalence (proportion of population). The target indicators CVD08 and CVD09
(coronary heart disease and all heart disease mortality) are reported as
Crude Rate (deaths per 100k). We keep both types and drop everything else.

```{r filter-type, cache=TRUE, cache.path="cache/", dependson="filter-suppress"}
is_crude_prev <- clean$DataValueType == "Crude Prevalence"
is_cvd_rate   <- (clean$QuestionID == "CVD08" | clean$QuestionID == "CVD09") &
                  clean$DataValueType == "Crude Rate"
clean <- clean[is_crude_prev | is_cvd_rate, ]
cat("Rows after filtering to relevant types:", nrow(clean), "\n")
cat("DataValueTypes remaining:", unique(clean$DataValueType), "\n")
```

### 2d. Align Year Ranges

Some records span multiple years (e.g., 2015-2019 rolling averages). We assign
each record to its `YearStart` as a single year key so all records can be
joined on a consistent time dimension.

```{r year-align, cache=TRUE, cache.path="cache/", dependson="filter-type"}
clean$Year <- clean$YearStart
clean <- clean[, setdiff(names(clean), c("YearStart", "YearEnd"))]
cat("Year range:", range(clean$Year), "\n")
```

### 2e. Filter to CVD & Risk Factor Topics

We keep Cardiovascular Disease records (which contain our target variables
CVD08/CVD09 — heart disease mortality rates — and CVD condition prevalence
features) and four established CVD risk factor topics: Tobacco, Diabetes,
Alcohol, and Nutrition/Physical Activity/Weight Status.

```{r filter-topics, cache=TRUE, cache.path="cache/", dependson="year-align"}
cvd_risk_topics <- c(
  "Cardiovascular Disease",
  "Tobacco",
  "Diabetes",
  "Alcohol",
  "Nutrition, Physical Activity, and Weight Status"
)
clean <- clean[is.element(clean$Topic, cvd_risk_topics), ]
cat("Rows after filtering to CVD + risk factor topics:", nrow(clean), "\n")
cat("Topics retained:\n")
print(table(clean$Topic))
```

### 2f. Pivot Long to Wide

Currently each row is one indicator measurement. We pivot to wide format so
each row represents a complete profile for a given state, year, and demographic
group. `QuestionID` becomes the column name; duplicate key combinations are
collapsed by taking the mean.

```{r pivot-wide, cache=TRUE, cache.path="cache/", dependson="filter-topics"}
sub_clean <- clean[, c("LocationAbbr", "Year", "Stratification1", "QuestionID", "DataValue")]
wide <- pivot_wider(
  sub_clean,
  names_from  = QuestionID,
  values_from = DataValue,
  values_fn   = mean
)
cat("Wide format dimensions:", dim(wide), "\n")
cat("Columns:", names(wide), "\n")
```

### 2g. Handle Missing Values After Pivot

After pivoting, cells with no recorded measurement become `NA`. We:
1. Drop rows where the target column (`CVD09`) is `NA` (this also removes the
   "Age 18-44" prevalence rows that have no matching "Age 0-44" mortality rows)
2. Drop any column where more than 50% of values are `NA` (indicator not
   broadly reported)
3. Impute remaining feature `NA`s with the column median

```{r missing-values, cache=TRUE, cache.path="cache/", dependson="pivot-wide"}
# 1. Drop rows with no target value first, before any column pruning
#    (CVD09 = Diseases of the heart mortality, Crude Rate per 100k)
wide <- wide[!is.na(wide[["CVD09"]]), ]
cat("Rows after dropping NA target rows:", nrow(wide), "\n")

# 2. Drop columns with >50% NA (target and key cols are protected by step 1)
na_fracs <- colMeans(is.na(wide))
wide <- wide[, na_fracs <= 0.5]
cat("Columns after dropping sparse cols:", ncol(wide), "\n")

# 3. Impute remaining NAs with column median
num_cols <- which(sapply(wide, is.numeric))
for (col in names(wide)[num_cols]) {
  na_idx <- is.na(wide[[col]])
  if (any(na_idx)) {
    wide[[col]][na_idx] <- median(wide[[col]], na.rm = TRUE)
  }
}
cat("Remaining NAs after imputation:", sum(is.na(wide)), "\n")
```

## 3. Target & Feature Separation

`CVD09` ("Diseases of the heart mortality among all people", Crude Rate per
100k) is the target variable. It captures broad cardiovascular mortality and
has 3 years of coverage (2019–2021). `CVD08` (coronary heart disease mortality
specifically) is retained as a feature alongside CVD condition prevalence
indicators (CVD01–CVD04) and the behavioral risk factor indicators.

```{r target-features, dependson="missing-values"}
cvd_target <- "CVD09"
feature_cols <- setdiff(names(wide), c("LocationAbbr", "Year", "Stratification1", cvd_target))

cat("Target variable:", cvd_target,
    "— Diseases of the heart mortality, Crude Rate per 100k\n")
cat("Number of feature columns:", length(feature_cols), "\n")
cat("Feature columns:\n")
print(feature_cols)

# Quick summary of target
summary(wide[[cvd_target]])
```

## 4. Train / Test Split

```{r train-test-split, dependson="missing-values"}
set.seed(160)
train_idx <- sample(nrow(wide), 0.8 * nrow(wide))
train <- wide[train_idx, ]
test  <- wide[-train_idx, ]

cat("Training rows:", nrow(train), "\n")
cat("Test rows:    ", nrow(test), "\n")
cat("Split check — total:", nrow(train) + nrow(test), "== wide rows:", nrow(wide), "\n")
```

## 5. Multiple Linear Regression

```{r mlr-model, dependson="train-test-split"}
# Build formula dynamically from feature columns
mlr_formula <- as.formula(paste("CVD09 ~", paste(feature_cols, collapse = " + ")))

# Fit OLS model on training data
mlr_model <- lm(mlr_formula, data = train)

# Summary of coefficients
summary(mlr_model)
```

```{r mlr-evaluate, dependson="mlr-model"}
# Predictions on test set
mlr_pred <- predict(mlr_model, newdata = test)

# Evaluation metrics
mlr_rmse <- sqrt(mean((mlr_pred - test$CVD09)^2))
mlr_r2   <- 1 - sum((mlr_pred - test$CVD09)^2) / sum((test$CVD09 - mean(test$CVD09))^2)

cat("--- Multiple Linear Regression ---\n")
cat("RMSE: ", round(mlr_rmse, 3), "\n")
cat("R²:   ", round(mlr_r2,   3), "\n")
```

```{r mlr-plots, dependson="mlr-evaluate"}
# Residuals vs Fitted
mlr_results <- data.frame(
  Fitted    = mlr_pred,
  Actual    = test$CVD09,
  Residuals = test$CVD09 - mlr_pred
)

# Residuals vs Fitted plot
ggplot(mlr_results, aes(x = Fitted, y = Residuals)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "MLR — Residuals vs Fitted",
    x     = "Fitted Values",
    y     = "Residuals"
  ) +
  theme_minimal()

# Actual vs Predicted plot
ggplot(mlr_results, aes(x = Actual, y = Fitted)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "MLR — Actual vs Predicted",
    x     = "Actual CVD09",
    y     = "Predicted CVD09"
  ) +
  theme_minimal()
```


## 6. Ridge Regression

```{r ridge-setup, message=FALSE, dependson="train-test-split"}
library(glmnet)

# glmnet requires matrices, not data frames
X_train <- as.matrix(train[, feature_cols])
y_train <- train$CVD09

X_test  <- as.matrix(test[, feature_cols])
y_test  <- test$CVD09
```

```{r ridge-cv, dependson="ridge-setup"}
# 10-fold cross-validation to find optimal lambda
set.seed(160)
cv_ridge <- cv.glmnet(
  X_train, y_train,
  alpha  = 0,       # alpha=0 means Ridge (alpha=1 would be Lasso)
  nfolds = 10
)

# Plot CV error across lambda values
plot(cv_ridge)
title("Ridge — Cross-Validation Error vs Lambda", line = 3)

cat("Optimal lambda (lambda.min): ", round(cv_ridge$lambda.min, 4), "\n")
cat("1-SE lambda (lambda.1se):    ", round(cv_ridge$lambda.1se, 4), "\n")
```

```{r ridge-evaluate, dependson="ridge-cv"}
# Predict on test set using optimal lambda
ridge_pred <- predict(cv_ridge, newx = X_test, s = "lambda.min")

# Evaluation metrics
ridge_rmse <- sqrt(mean((ridge_pred - y_test)^2))
ridge_r2   <- 1 - sum((ridge_pred - y_test)^2) / sum((y_test - mean(y_test))^2)

cat("--- Ridge Regression ---\n")
cat("RMSE: ", round(ridge_rmse, 3), "\n")
cat("R²:   ", round(ridge_r2,   3), "\n")
```

```{r ridge-coefficients, dependson="ridge-cv"}
# Extract and display shrunken coefficients
ridge_coefs <- coef(cv_ridge, s = "lambda.min")
ridge_coef_df <- data.frame(
  Feature     = rownames(ridge_coefs),
  Coefficient = as.vector(ridge_coefs)
) %>%
  filter(Feature != "(Intercept)") %>%
  arrange(desc(abs(Coefficient)))

print(ridge_coef_df)

# Plot top 15 coefficients by magnitude
ridge_coef_df %>%
  slice_head(n = 15) %>%
  ggplot(aes(x = reorder(Feature, abs(Coefficient)), y = Coefficient)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Ridge — Top 15 Coefficients by Magnitude",
    x     = "Feature",
    y     = "Coefficient"
  ) +
  theme_minimal()
```

```{r ridge-plots, dependson="ridge-evaluate"}
ridge_results <- data.frame(
  Fitted    = as.vector(ridge_pred),
  Actual    = y_test,
  Residuals = y_test - as.vector(ridge_pred)
)

# Residuals vs Fitted
ggplot(ridge_results, aes(x = Fitted, y = Residuals)) +
  geom_point(alpha = 0.6, color = "darkorange") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Ridge — Residuals vs Fitted",
    x     = "Fitted Values",
    y     = "Residuals"
  ) +
  theme_minimal()

# Actual vs Predicted
ggplot(ridge_results, aes(x = Actual, y = Fitted)) +
  geom_point(alpha = 0.6, color = "darkorange") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Ridge — Actual vs Predicted",
    x     = "Actual CVD09",
    y     = "Predicted CVD09"
  ) +
  theme_minimal()
```


## 7. Model Comparison

```{r model-comparison, dependson=c("mlr-evaluate", "ridge-evaluate")}
comparison <- data.frame(
  Model = c("Multiple Linear Regression", "Ridge Regression"),
  RMSE  = c(mlr_rmse, ridge_rmse),
  R2    = c(mlr_r2,   ridge_r2)
)

print(comparison)

# Bar chart comparison
comparison_long <- comparison %>%
  pivot_longer(cols = c(RMSE, R2), names_to = "Metric", values_to = "Value")

ggplot(comparison_long, aes(x = Model, y = Value, fill = Model)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ Metric, scales = "free_y") +
  labs(
    title = "Model Comparison — MLR vs Ridge",
    x     = NULL,
    y     = "Value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```


## 8. Random Forest — Mortality Range Classification

CVD09 mortality rates are split into five equal-frequency quintile bands
(Very Low through Very High) using the full dataset distribution. The random
forest then predicts which band a state/year/demographic row falls into.

### 8a. Define Quintile Ranges

```{r rf-setup, dependson="train-test-split"}
cvd09_breaks <- quantile(wide$CVD09, probs = seq(0, 1, by = 0.2), na.rm = TRUE)
range_labels <- c("Very Low", "Low", "Medium", "High", "Very High")

train_rf <- train
test_rf  <- test
train_rf$CVD09_range <- cut(train_rf$CVD09, breaks = cvd09_breaks,
                             labels = range_labels, include.lowest = TRUE)
test_rf$CVD09_range  <- cut(test_rf$CVD09,  breaks = cvd09_breaks,
                             labels = range_labels, include.lowest = TRUE)

cat("Quintile break points (deaths per 100k):\n")
print(round(cvd09_breaks, 1))
cat("\nTraining class distribution:\n")
print(table(train_rf$CVD09_range))
cat("\nTest class distribution:\n")
print(table(test_rf$CVD09_range))
```

### 8b. Fit Random Forest

```{r rf-model, message=FALSE, dependson="rf-setup"}
library(randomForest)
set.seed(160)
rf_formula <- as.formula(paste("CVD09_range ~", paste(feature_cols, collapse = " + ")))
rf_model <- randomForest(rf_formula, data = train_rf, ntree = 500, importance = TRUE)
print(rf_model)
```

### 8c. Evaluate

```{r rf-evaluate, dependson="rf-model"}
rf_pred     <- predict(rf_model, newdata = test_rf)
conf_mat_rf <- table(Predicted = rf_pred, Actual = test_rf$CVD09_range)
rf_accuracy <- mean(rf_pred == test_rf$CVD09_range)

cat("Confusion Matrix:\n")
print(conf_mat_rf)
cat("\nOverall Accuracy:", round(rf_accuracy, 3), "\n")

conf_df <- as.data.frame(conf_mat_rf)
ggplot(conf_df, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), size = 6, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "forestgreen") +
  labs(
    title = "Random Forest — Confusion Matrix",
    x     = "Actual Range",
    y     = "Predicted Range"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

### 8d. Variable Importance

```{r rf-importance, dependson="rf-model"}
rf_imp_raw <- importance(rf_model, type = 1)
rf_imp_df  <- data.frame(
  Feature    = rownames(rf_imp_raw),
  Importance = rf_imp_raw[, 1]
)
rf_imp_df <- rf_imp_df[order(rf_imp_df$Importance, decreasing = TRUE), ]

ggplot(rf_imp_df[1:min(15, nrow(rf_imp_df)), ],
       aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = "forestgreen") +
  coord_flip() +
  labs(
    title = "Random Forest — Top 15 Features (Mean Decrease in Accuracy)",
    x     = "Feature",
    y     = "Mean Decrease in Accuracy"
  ) +
  theme_minimal()
```


## 9. Feature Importance — All Models

Each model measures feature importance on a different scale, so we normalize
each to [0, 1] before comparing. Sources:

- **MLR**: absolute t-statistic (coefficient divided by its standard error)
- **Ridge**: absolute shrunken coefficient (from Section 6)
- **Logistic**: absolute z-statistic (`log_model` must be available — run `LogisticRegression.Rmd` first in the same session)
- **Random Forest**: mean decrease in accuracy when a feature is permuted (from Section 8)

Sections 6 and 8 already show Ridge and RF importance individually; this
section adds MLR and Logistic plots, then combines all four into a consensus
ranking.

### 9a. MLR Feature Importance

```{r importance-mlr, dependson="mlr-model"}
mlr_coef_mat <- summary(mlr_model)$coefficients
mlr_imp_df <- data.frame(
  Feature    = rownames(mlr_coef_mat)[-1],
  Importance = abs(mlr_coef_mat[-1, "t value"])
)
mlr_imp_df <- mlr_imp_df[order(mlr_imp_df$Importance, decreasing = TRUE), ]

ggplot(mlr_imp_df[1:min(15, nrow(mlr_imp_df)), ],
       aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "MLR — Top 15 Features by Absolute t-Statistic",
    x     = "Feature",
    y     = "Absolute t-Statistic"
  ) +
  theme_minimal()
```

### 9b. Logistic Regression Feature Importance

```{r importance-logistic}
if (!exists("log_model")) {
  message("log_model not found — run LogisticRegression.Rmd in this session first. Skipping.")
  log_imp_df <- data.frame(Feature = character(0), Importance = numeric(0))
} else {
  log_coef_mat <- summary(log_model)$coefficients
  log_imp_df <- data.frame(
    Feature    = rownames(log_coef_mat)[-1],
    Importance = abs(log_coef_mat[-1, "z value"])
  )
  log_imp_df <- log_imp_df[!is.na(log_imp_df$Importance), ]
  log_imp_df <- log_imp_df[order(log_imp_df$Importance, decreasing = TRUE), ]

  print(
    ggplot(log_imp_df[1:min(15, nrow(log_imp_df)), ],
           aes(x = reorder(Feature, Importance), y = Importance)) +
      geom_col(fill = "darkorange") +
      coord_flip() +
      labs(
        title = "Logistic Regression — Top 15 Features by Absolute z-Statistic",
        x     = "Feature",
        y     = "Absolute z-Statistic"
      ) +
      theme_minimal()
  )
}
```

### 9c. Consensus Ranking

```{r importance-consensus, dependson=c("mlr-model", "ridge-cv", "rf-model")}
normalize_imp <- function(x) {
  rng <- range(x, na.rm = TRUE)
  if (rng[1] == rng[2]) return(rep(0, length(x)))
  (x - rng[1]) / (rng[2] - rng[1])
}

mlr_norm <- data.frame(
  Feature = mlr_imp_df$Feature,
  MLR     = normalize_imp(mlr_imp_df$Importance)
)

ridge_coef_vec <- as.vector(coef(cv_ridge, s = "lambda.min"))[-1]
ridge_norm <- data.frame(
  Feature = rownames(coef(cv_ridge, s = "lambda.min"))[-1],
  Ridge   = normalize_imp(abs(ridge_coef_vec))
)

if (nrow(log_imp_df) > 0) {
  log_norm <- data.frame(
    Feature  = log_imp_df$Feature,
    Logistic = normalize_imp(log_imp_df$Importance)
  )
} else {
  log_norm <- data.frame(Feature = character(0), Logistic = numeric(0))
}

rf_norm <- data.frame(
  Feature = rf_imp_df$Feature,
  RF      = normalize_imp(rf_imp_df$Importance)
)

imp_all <- merge(mlr_norm,  ridge_norm, by = "Feature", all = TRUE)
imp_all <- merge(imp_all,   log_norm,   by = "Feature", all = TRUE)
imp_all <- merge(imp_all,   rf_norm,    by = "Feature", all = TRUE)

imp_all$Consensus <- rowMeans(
  imp_all[, c("MLR", "Ridge", "Logistic", "RF")], na.rm = TRUE
)
imp_all <- imp_all[order(imp_all$Consensus, decreasing = TRUE), ]

cat("Top 15 features by consensus importance:\n")
print(imp_all[1:min(15, nrow(imp_all)), ])

top15 <- imp_all[1:min(15, nrow(imp_all)), ]
ggplot(top15, aes(x = reorder(Feature, Consensus), y = Consensus)) +
  geom_col(fill = "purple4") +
  coord_flip() +
  labs(
    title = "Consensus Feature Importance — Average Across All 4 Models",
    x     = "Feature",
    y     = "Mean Normalized Importance (0 to 1)"
  ) +
  theme_minimal()
```

