---
title: "CDC Chronic Disease Indicators - CVD Prevalence Project"
author: "Lenio, Novick, Tanveer, Thompson"
date: "12 April 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
```

## 1. Load & Cache

The raw dataset is large (~300k rows). We load it once and cache it so
subsequent knits skip the file read entirely.

```{r load, cache=TRUE, cache.path="cache/"}
raw <- read.csv("U.S._Chronic_Disease_Indicators.csv", stringsAsFactors = FALSE)
cat("Loaded:", nrow(raw), "rows,", ncol(raw), "columns\n")
```

## 2. Preprocessing

### 2a. Drop Sparse & Unnecessary Columns

Keep only the columns needed for modeling. The secondary stratification fields
(`Stratification2/3`, `ResponseID`, ID columns, etc.) are empty or near-empty
across the full dataset and are dropped here.

```{r drop-columns, cache=TRUE, cache.path="cache/", dependson="load"}
clean <- raw %>%
  select(
    YearStart, YearEnd, LocationAbbr,
    Topic, Question, QuestionID,
    DataValueType, DataValue,
    LowConfidenceLimit, HighConfidenceLimit,
    StratificationCategory1, Stratification1
  )
cat("Columns kept:", ncol(clean), "\n")
cat("Rows:", nrow(clean), "\n")
```

### 2b. Remove Suppressed Cells

Some cells have no `DataValue` because the sample was too small to report.
These are removed rather than imputed, as they represent structural missingness.

```{r filter-suppress, cache=TRUE, cache.path="cache/", dependson="drop-columns"}
clean <- clean %>% filter(!is.na(DataValue))
cat("Rows after removing suppressed values:", nrow(clean), "\n")
cat("Remaining NAs in DataValue:", sum(is.na(clean$DataValue)), "\n")
```

### 2c. Filter to Crude Prevalence

The dataset mixes rates, counts, percentages, and means across different
`DataValueType` values. Filtering to `"Crude Prevalence"` keeps only
percentage-of-population measures, which are directly comparable across
indicators and align with our target variable definition.

```{r filter-type, cache=TRUE, cache.path="cache/", dependson="filter-suppress"}
clean <- clean %>% filter(DataValueType == "Crude Prevalence")
cat("Rows after filtering to Crude Prevalence:", nrow(clean), "\n")
cat("DataValueTypes remaining:", unique(clean$DataValueType), "\n")
```

### 2d. Align Year Ranges

Some records span multiple years (e.g., 2015–2019 rolling averages). We assign
each record to its `YearStart` as a single year key so all records can be
joined on a consistent time dimension.

```{r year-align, cache=TRUE, cache.path="cache/", dependson="filter-type"}
clean <- clean %>%
  mutate(Year = YearStart) %>%
  select(-YearStart, -YearEnd)
cat("Year range:", range(clean$Year), "\n")
```

### 2e. Filter to CVD & Risk Factor Topics

We keep Cardiovascular Disease records (which contain our target variable,
CVD01 — high blood pressure prevalence) and four established CVD risk factor
topics: Tobacco, Diabetes, Alcohol, and Nutrition/Physical Activity/Weight
Status.

```{r filter-topics, cache=TRUE, cache.path="cache/", dependson="year-align"}
cvd_risk_topics <- c(
  "Cardiovascular Disease",
  "Tobacco",
  "Diabetes",
  "Alcohol",
  "Nutrition, Physical Activity, and Weight Status"
)
clean <- clean %>% filter(Topic %in% cvd_risk_topics)
cat("Rows after filtering to CVD + risk factor topics:", nrow(clean), "\n")
cat("Topics retained:\n")
print(table(clean$Topic))
```

### 2f. Pivot Long to Wide

Currently each row is one indicator measurement. We pivot to wide format so
each row represents a complete profile for a given state, year, and demographic
group. `QuestionID` becomes the column name; duplicate key combinations are
collapsed by taking the mean.

```{r pivot-wide, cache=TRUE, cache.path="cache/", dependson="filter-topics"}
wide <- clean %>%
  select(LocationAbbr, Year, Stratification1, QuestionID, DataValue) %>%
  pivot_wider(
    names_from  = QuestionID,
    values_from = DataValue,
    values_fn   = mean
  )
cat("Wide format dimensions:", dim(wide), "\n")
cat("Columns:", names(wide), "\n")
```

### 2g. Handle Missing Values After Pivot

After pivoting, cells with no recorded measurement become `NA`. We:
1. Drop any column where more than 50% of values are `NA` (indicator not
   broadly reported)
2. Drop rows where the target column (`CVD01`) is `NA`
3. Impute remaining feature `NA`s with the column median

```{r missing-values, cache=TRUE, cache.path="cache/", dependson="pivot-wide"}
# 1. Drop rows with no target value first, before any column pruning
#    (CVD01 = High blood pressure prevalence)
wide <- wide %>% filter(!is.na(.data[["CVD01"]]))
cat("Rows after dropping NA target rows:", nrow(wide), "\n")

# 2. Drop columns with >50% NA (target and key cols are protected by step 1)
wide <- wide %>% select(where(~ mean(is.na(.)) <= 0.5))
cat("Columns after dropping sparse cols:", ncol(wide), "\n")

# 3. Impute remaining NAs with column median
wide <- wide %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), median(., na.rm = TRUE), .)))
cat("Remaining NAs after imputation:", sum(is.na(wide)), "\n")
```

## 3. Target & Feature Separation

`CVD01` ("High blood pressure among adults", Crude Prevalence %) is used as
our CVD prevalence proxy — it is the most broadly reported CVD prevalence
measure in the dataset. The remaining numeric columns from CVD risk factor
topics serve as predictive features.

```{r target-features, dependson="missing-values"}
cvd_target <- "CVD01"
feature_cols <- setdiff(names(wide), c("LocationAbbr", "Year", "Stratification1", cvd_target))

cat("Target variable:", cvd_target,
    "— High blood pressure among adults (Crude Prevalence %)\n")
cat("Number of feature columns:", length(feature_cols), "\n")
cat("Feature columns:\n")
print(feature_cols)

# Quick summary of target
summary(wide[[cvd_target]])
```

## 4. Train / Test Split

```{r train-test-split, dependson="missing-values"}
set.seed(160)
train_idx <- sample(nrow(wide), 0.8 * nrow(wide))
train <- wide[train_idx, ]
test  <- wide[-train_idx, ]

cat("Training rows:", nrow(train), "\n")
cat("Test rows:    ", nrow(test), "\n")
cat("Split check — total:", nrow(train) + nrow(test), "== wide rows:", nrow(wide), "\n")
```