---
title: "CDC Chronic Disease Indicators - CVD Prevalence Project"
author: "Lenio, Novick, Tanveer, Thompson"
date: "12 April 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
```

## 1. Load & Cache

The raw dataset is large (~300k rows). We load it once and cache it so
subsequent knits skip the file read entirely.

```{r load, cache=TRUE, cache.path="cache/"}
raw <- read.csv("U.S._Chronic_Disease_Indicators.csv", stringsAsFactors = FALSE)
cat("Loaded:", nrow(raw), "rows,", ncol(raw), "columns\n")
```

## 2. Preprocessing

### 2a. Drop Sparse & Unnecessary Columns

Keep only the columns needed for modeling. The secondary stratification fields
(`Stratification2/3`, `ResponseID`, ID columns, etc.) are empty or near-empty
across the full dataset and are dropped here.

```{r drop-columns, cache=TRUE, cache.path="cache/", dependson="load"}
clean <- raw %>%
  select(
    YearStart, YearEnd, LocationAbbr,
    Topic, Question, QuestionID,
    DataValueType, DataValue,
    LowConfidenceLimit, HighConfidenceLimit,
    StratificationCategory1, Stratification1
  )
cat("Columns kept:", ncol(clean), "\n")
cat("Rows:", nrow(clean), "\n")
```

### 2b. Remove Suppressed Cells

Some cells have no `DataValue` because the sample was too small to report.
These are removed rather than imputed, as they represent structural missingness.

```{r filter-suppress, cache=TRUE, cache.path="cache/", dependson="drop-columns"}
clean <- clean %>% filter(!is.na(DataValue))
cat("Rows after removing suppressed values:", nrow(clean), "\n")
cat("Remaining NAs in DataValue:", sum(is.na(clean$DataValue)), "\n")
```

### 2c. Filter to Relevant Data Value Types

Feature indicators (tobacco, diabetes, obesity, CVD conditions) are reported as
`"Crude Prevalence"` (% of population). The target indicators CVD08 and CVD09
(coronary heart disease and all heart disease mortality) are reported as
`"Crude Rate"` (deaths per 100k). We keep both types and drop everything else.

```{r filter-type, cache=TRUE, cache.path="cache/", dependson="filter-suppress"}
clean <- clean %>%
  filter(
    DataValueType == "Crude Prevalence" |
    (QuestionID %in% c("CVD08", "CVD09") & DataValueType == "Crude Rate")
  )
cat("Rows after filtering to relevant types:", nrow(clean), "\n")
cat("DataValueTypes remaining:", unique(clean$DataValueType), "\n")
```

### 2d. Align Year Ranges

Some records span multiple years (e.g., 2015–2019 rolling averages). We assign
each record to its `YearStart` as a single year key so all records can be
joined on a consistent time dimension.

```{r year-align, cache=TRUE, cache.path="cache/", dependson="filter-type"}
clean <- clean %>%
  mutate(Year = YearStart) %>%
  select(-YearStart, -YearEnd)
cat("Year range:", range(clean$Year), "\n")
```

### 2e. Filter to CVD & Risk Factor Topics

We keep Cardiovascular Disease records (which contain our target variables
CVD08/CVD09 — heart disease mortality rates — and CVD condition prevalence
features) and four established CVD risk factor topics: Tobacco, Diabetes,
Alcohol, and Nutrition/Physical Activity/Weight Status.

```{r filter-topics, cache=TRUE, cache.path="cache/", dependson="year-align"}
cvd_risk_topics <- c(
  "Cardiovascular Disease",
  "Tobacco",
  "Diabetes",
  "Alcohol",
  "Nutrition, Physical Activity, and Weight Status"
)
clean <- clean %>% filter(Topic %in% cvd_risk_topics)
cat("Rows after filtering to CVD + risk factor topics:", nrow(clean), "\n")
cat("Topics retained:\n")
print(table(clean$Topic))
```

### 2f. Pivot Long to Wide

Currently each row is one indicator measurement. We pivot to wide format so
each row represents a complete profile for a given state, year, and demographic
group. `QuestionID` becomes the column name; duplicate key combinations are
collapsed by taking the mean.

```{r pivot-wide, cache=TRUE, cache.path="cache/", dependson="filter-topics"}
wide <- clean %>%
  select(LocationAbbr, Year, Stratification1, QuestionID, DataValue) %>%
  pivot_wider(
    names_from  = QuestionID,
    values_from = DataValue,
    values_fn   = mean
  )
cat("Wide format dimensions:", dim(wide), "\n")
cat("Columns:", names(wide), "\n")
```

### 2g. Handle Missing Values After Pivot

After pivoting, cells with no recorded measurement become `NA`. We:
1. Drop rows where the target column (`CVD09`) is `NA` (this also removes the
   "Age 18-44" prevalence rows that have no matching "Age 0-44" mortality rows)
2. Drop any column where more than 50% of values are `NA` (indicator not
   broadly reported)
3. Impute remaining feature `NA`s with the column median

```{r missing-values, cache=TRUE, cache.path="cache/", dependson="pivot-wide"}
# 1. Drop rows with no target value first, before any column pruning
#    (CVD09 = Diseases of the heart mortality, Crude Rate per 100k)
wide <- wide %>% filter(!is.na(.data[["CVD09"]]))
cat("Rows after dropping NA target rows:", nrow(wide), "\n")

# 2. Drop columns with >50% NA (target and key cols are protected by step 1)
wide <- wide %>% select(where(~ mean(is.na(.)) <= 0.5))
cat("Columns after dropping sparse cols:", ncol(wide), "\n")

# 3. Impute remaining NAs with column median
wide <- wide %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), median(., na.rm = TRUE), .)))
cat("Remaining NAs after imputation:", sum(is.na(wide)), "\n")
```

## 3. Target & Feature Separation

`CVD09` ("Diseases of the heart mortality among all people", Crude Rate per
100k) is the target variable. It captures broad cardiovascular mortality and
has 3 years of coverage (2019–2021). `CVD08` (coronary heart disease mortality
specifically) is retained as a feature alongside CVD condition prevalence
indicators (CVD01–CVD04) and the behavioral risk factor indicators.

```{r target-features, dependson="missing-values"}
cvd_target <- "CVD09"
feature_cols <- setdiff(names(wide), c("LocationAbbr", "Year", "Stratification1", cvd_target))

cat("Target variable:", cvd_target,
    "— Diseases of the heart mortality, Crude Rate per 100k\n")
cat("Number of feature columns:", length(feature_cols), "\n")
cat("Feature columns:\n")
print(feature_cols)

# Quick summary of target
summary(wide[[cvd_target]])
```

## 4. Train / Test Split

```{r train-test-split, dependson="missing-values"}
set.seed(160)
train_idx <- sample(nrow(wide), 0.8 * nrow(wide))
train <- wide[train_idx, ]
test  <- wide[-train_idx, ]

cat("Training rows:", nrow(train), "\n")
cat("Test rows:    ", nrow(test), "\n")
cat("Split check — total:", nrow(train) + nrow(test), "== wide rows:", nrow(wide), "\n")
```

## 5. Multiple Linear Regression

```{r mlr-model, dependson="train-test-split"}
# Build formula dynamically from feature columns
mlr_formula <- as.formula(paste("CVD09 ~", paste(feature_cols, collapse = " + ")))

# Fit OLS model on training data
mlr_model <- lm(mlr_formula, data = train)

# Summary of coefficients
summary(mlr_model)
```

```{r mlr-evaluate, dependson="mlr-model"}
# Predictions on test set
mlr_pred <- predict(mlr_model, newdata = test)

# Evaluation metrics
mlr_rmse <- sqrt(mean((mlr_pred - test$CVD09)^2))
mlr_r2   <- 1 - sum((mlr_pred - test$CVD09)^2) / sum((test$CVD09 - mean(test$CVD09))^2)

cat("--- Multiple Linear Regression ---\n")
cat("RMSE: ", round(mlr_rmse, 3), "\n")
cat("R²:   ", round(mlr_r2,   3), "\n")
```

```{r mlr-plots, dependson="mlr-evaluate"}
# Residuals vs Fitted
mlr_results <- data.frame(
  Fitted    = mlr_pred,
  Actual    = test$CVD09,
  Residuals = test$CVD09 - mlr_pred
)

# Residuals vs Fitted plot
ggplot(mlr_results, aes(x = Fitted, y = Residuals)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "MLR — Residuals vs Fitted",
    x     = "Fitted Values",
    y     = "Residuals"
  ) +
  theme_minimal()

# Actual vs Predicted plot
ggplot(mlr_results, aes(x = Actual, y = Fitted)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "MLR — Actual vs Predicted",
    x     = "Actual CVD09",
    y     = "Predicted CVD09"
  ) +
  theme_minimal()
```


## 6. Ridge Regression

```{r ridge-setup, message=FALSE, dependson="train-test-split"}
library(glmnet)

# glmnet requires matrices, not data frames
X_train <- as.matrix(train[, feature_cols])
y_train <- train$CVD09

X_test  <- as.matrix(test[, feature_cols])
y_test  <- test$CVD09
```

```{r ridge-cv, dependson="ridge-setup"}
# 10-fold cross-validation to find optimal lambda
set.seed(160)
cv_ridge <- cv.glmnet(
  X_train, y_train,
  alpha  = 0,       # alpha=0 means Ridge (alpha=1 would be Lasso)
  nfolds = 10
)

# Plot CV error across lambda values
plot(cv_ridge)
title("Ridge — Cross-Validation Error vs Lambda", line = 3)

cat("Optimal lambda (lambda.min): ", round(cv_ridge$lambda.min, 4), "\n")
cat("1-SE lambda (lambda.1se):    ", round(cv_ridge$lambda.1se, 4), "\n")
```

```{r ridge-evaluate, dependson="ridge-cv"}
# Predict on test set using optimal lambda
ridge_pred <- predict(cv_ridge, newx = X_test, s = "lambda.min")

# Evaluation metrics
ridge_rmse <- sqrt(mean((ridge_pred - y_test)^2))
ridge_r2   <- 1 - sum((ridge_pred - y_test)^2) / sum((y_test - mean(y_test))^2)

cat("--- Ridge Regression ---\n")
cat("RMSE: ", round(ridge_rmse, 3), "\n")
cat("R²:   ", round(ridge_r2,   3), "\n")
```

```{r ridge-coefficients, dependson="ridge-cv"}
# Extract and display shrunken coefficients
ridge_coefs <- coef(cv_ridge, s = "lambda.min")
ridge_coef_df <- data.frame(
  Feature     = rownames(ridge_coefs),
  Coefficient = as.vector(ridge_coefs)
) %>%
  filter(Feature != "(Intercept)") %>%
  arrange(desc(abs(Coefficient)))

print(ridge_coef_df)

# Plot top 15 coefficients by magnitude
ridge_coef_df %>%
  slice_head(n = 15) %>%
  ggplot(aes(x = reorder(Feature, abs(Coefficient)), y = Coefficient)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Ridge — Top 15 Coefficients by Magnitude",
    x     = "Feature",
    y     = "Coefficient"
  ) +
  theme_minimal()
```

```{r ridge-plots, dependson="ridge-evaluate"}
ridge_results <- data.frame(
  Fitted    = as.vector(ridge_pred),
  Actual    = y_test,
  Residuals = y_test - as.vector(ridge_pred)
)

# Residuals vs Fitted
ggplot(ridge_results, aes(x = Fitted, y = Residuals)) +
  geom_point(alpha = 0.6, color = "darkorange") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Ridge — Residuals vs Fitted",
    x     = "Fitted Values",
    y     = "Residuals"
  ) +
  theme_minimal()

# Actual vs Predicted
ggplot(ridge_results, aes(x = Actual, y = Fitted)) +
  geom_point(alpha = 0.6, color = "darkorange") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Ridge — Actual vs Predicted",
    x     = "Actual CVD09",
    y     = "Predicted CVD09"
  ) +
  theme_minimal()
```


## 7. Model Comparison

```{r model-comparison, dependson=c("mlr-evaluate", "ridge-evaluate")}
comparison <- data.frame(
  Model = c("Multiple Linear Regression", "Ridge Regression"),
  RMSE  = c(mlr_rmse, ridge_rmse),
  R2    = c(mlr_r2,   ridge_r2)
)

print(comparison)

# Bar chart comparison
comparison_long <- comparison %>%
  pivot_longer(cols = c(RMSE, R2), names_to = "Metric", values_to = "Value")

ggplot(comparison_long, aes(x = Model, y = Value, fill = Model)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ Metric, scales = "free_y") +
  labs(
    title = "Model Comparison — MLR vs Ridge",
    x     = NULL,
    y     = "Value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```
