---
title: "Stratification — Adding Location and Demographics"
author: "Lenio, Novick, Tanveer, Thompson"
date: "12 April 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The baseline models treat every row as an anonymous observation — they see
indicator values but not which state or demographic group the row represents.
This notebook first visualizes how CVD09 mortality differs across states and
demographic groups, then adds `LocationAbbr` and `Stratification1` as explicit
model inputs (with CVD08 excluded) to see whether knowing where and who
improves predictions.

**Requires:** Run `cse160_project.Rmd` and `CVD08_Ablation.Rmd` first in the
same session so that `wide`, `train`, `test`, `train_rf`, `test_rf`,
`feature_cols_no_cvd08`, `mlr_rmse`, `mlr_r2`, `ridge_rmse`, `ridge_r2`,
`rf_accuracy`, `y_train`, and `y_test` are all available.

```{r libraries, message=FALSE}
library(ggplot2)
library(glmnet)
library(randomForest)
```

## CVD09 by Demographic Group

```{r strat-explore-demo}
cvd09_by_demo <- aggregate(CVD09 ~ Stratification1, data = wide,
                            FUN = mean, na.rm = TRUE)
cvd09_by_demo <- cvd09_by_demo[order(cvd09_by_demo$CVD09, decreasing = TRUE), ]

ggplot(cvd09_by_demo, aes(x = reorder(Stratification1, CVD09), y = CVD09)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Mean CVD09 Mortality by Demographic Group",
    x     = NULL,
    y     = "Mean CVD09 (deaths per 100k)"
  ) +
  theme_minimal()
```

## CVD09 by State

```{r strat-explore-state}
cvd09_by_state <- aggregate(CVD09 ~ LocationAbbr, data = wide,
                              FUN = mean, na.rm = TRUE)
cvd09_by_state <- cvd09_by_state[order(cvd09_by_state$CVD09, decreasing = TRUE), ]

ggplot(cvd09_by_state, aes(x = reorder(LocationAbbr, CVD09), y = CVD09)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Mean CVD09 Mortality by State",
    x     = NULL,
    y     = "Mean CVD09 (deaths per 100k)"
  ) +
  theme_minimal()
```

## Encode Stratification Variables

`lm` and `randomForest` accept factors directly. For Ridge (`glmnet`) we need
a numeric matrix, so we use `model.matrix` with factor levels fixed to the
full dataset to guarantee train and test produce identical columns.

```{r strat-encode}
loc_levels   <- sort(unique(wide$LocationAbbr))
strat_levels <- sort(unique(wide$Stratification1))

train_f <- train
test_f  <- test
train_f$LocationAbbr    <- factor(train_f$LocationAbbr,    levels = loc_levels)
train_f$Stratification1 <- factor(train_f$Stratification1, levels = strat_levels)
test_f$LocationAbbr     <- factor(test_f$LocationAbbr,     levels = loc_levels)
test_f$Stratification1  <- factor(test_f$Stratification1,  levels = strat_levels)

mm_formula    <- as.formula(paste("~", paste(c(feature_cols_no_cvd08, "LocationAbbr", "Stratification1"),
                                              collapse = " + ")))
X_train_strat <- model.matrix(mm_formula, data = train_f)[, -1]
X_test_strat  <- model.matrix(mm_formula, data = test_f)[, -1]

train_rf_f <- train_rf
test_rf_f  <- test_rf
train_rf_f$LocationAbbr    <- factor(train_rf_f$LocationAbbr,    levels = loc_levels)
train_rf_f$Stratification1 <- factor(train_rf_f$Stratification1, levels = strat_levels)
test_rf_f$LocationAbbr     <- factor(test_rf_f$LocationAbbr,     levels = loc_levels)
test_rf_f$Stratification1  <- factor(test_rf_f$Stratification1,  levels = strat_levels)

cat("Dummy columns added for Ridge:", ncol(X_train_strat) - length(feature_cols_no_cvd08), "\n")
cat("Total Ridge input columns:    ", ncol(X_train_strat), "\n")
```

## MLR with Stratification

```{r strat-mlr}
mlr_strat_formula <- as.formula(paste(
  "CVD09 ~",
  paste(c(feature_cols_no_cvd08, "LocationAbbr", "Stratification1"), collapse = " + ")
))
mlr_strat_model <- lm(mlr_strat_formula, data = train_f)
mlr_strat_pred  <- predict(mlr_strat_model, newdata = test_f)

mlr_strat_rmse <- sqrt(mean((mlr_strat_pred - test_f$CVD09)^2))
mlr_strat_r2   <- 1 - sum((mlr_strat_pred - test_f$CVD09)^2) /
                      sum((test_f$CVD09 - mean(test_f$CVD09))^2)

cat("MLR baseline         — RMSE:", round(mlr_rmse,       3), "| R2:", round(mlr_r2,       3), "\n")
cat("MLR + stratification — RMSE:", round(mlr_strat_rmse, 3), "| R2:", round(mlr_strat_r2, 3), "\n")
```

## Ridge with Stratification

```{r strat-ridge}
set.seed(160)
cv_ridge_strat   <- cv.glmnet(X_train_strat, y_train, alpha = 0, nfolds = 10)
ridge_strat_pred <- predict(cv_ridge_strat, newx = X_test_strat, s = "lambda.min")

ridge_strat_rmse <- sqrt(mean((ridge_strat_pred - y_test)^2))
ridge_strat_r2   <- 1 - sum((ridge_strat_pred - y_test)^2) /
                        sum((y_test - mean(y_test))^2)

cat("Ridge baseline         — RMSE:", round(ridge_rmse,       3), "| R2:", round(ridge_r2,       3), "\n")
cat("Ridge + stratification — RMSE:", round(ridge_strat_rmse, 3), "| R2:", round(ridge_strat_r2, 3), "\n")
```

## Random Forest with Stratification

```{r strat-rf}
rf_strat_formula <- as.formula(paste(
  "CVD09_range ~",
  paste(c(feature_cols_no_cvd08, "LocationAbbr", "Stratification1"), collapse = " + ")
))
set.seed(160)
rf_strat_model    <- randomForest(rf_strat_formula, data = train_rf_f, ntree = 500)
rf_strat_pred     <- predict(rf_strat_model, newdata = test_rf_f)
rf_strat_accuracy <- mean(rf_strat_pred == test_rf_f$CVD09_range)

cat("RF baseline          — Accuracy:", round(rf_accuracy,       3), "\n")
cat("RF + stratification  — Accuracy:", round(rf_strat_accuracy, 3), "\n")
```

## Comparison

```{r strat-comparison}
reg_strat_comp <- data.frame(
  Model = c("MLR", "MLR + Strat", "Ridge", "Ridge + Strat"),
  RMSE  = c(mlr_rmse,   mlr_strat_rmse,   ridge_rmse,   ridge_strat_rmse),
  R2    = c(mlr_r2,     mlr_strat_r2,     ridge_r2,     ridge_strat_r2)
)
cat("Regression models:\n")
print(reg_strat_comp)

cat("\nRandom Forest:\n")
cat("  Baseline:        ", round(rf_accuracy,       3), "\n")
cat("  + Stratification:", round(rf_strat_accuracy, 3), "\n")

ggplot(reg_strat_comp, aes(x = Model, y = RMSE, fill = Model)) +
  geom_col(show.legend = FALSE) +
  labs(
    title = "RMSE — Baseline vs With Stratification",
    x     = NULL,
    y     = "RMSE (deaths per 100k)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

ggplot(reg_strat_comp, aes(x = Model, y = R2, fill = Model)) +
  geom_col(show.legend = FALSE) +
  labs(
    title = "R-Squared — Baseline vs With Stratification",
    x     = NULL,
    y     = "R-Squared"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

rf_strat_acc_df <- data.frame(
  Model    = c("RF Baseline", "RF + Stratification"),
  Accuracy = c(rf_accuracy, rf_strat_accuracy)
)
ggplot(rf_strat_acc_df, aes(x = Model, y = Accuracy, fill = Model)) +
  geom_col(show.legend = FALSE) +
  ylim(0, 1) +
  labs(
    title = "Random Forest Accuracy — Baseline vs With Stratification",
    x     = NULL,
    y     = "Accuracy"
  ) +
  theme_minimal()
```
