---
title: "CVD08 Ablation — Models Without CVD08"
author: "Lenio, Novick, Tanveer, Thompson"
date: "12 April 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

CVD08 (coronary heart disease mortality) is the strongest predictor across all
models, but it is reported on the same timeline as CVD09 and comes from the
same data source. In a real deployment scenario a user who already has CVD08
would likely already have CVD09, making the feature nearly circular. This
notebook reruns all three models with CVD08 removed to show how much predictive
power actually comes from the independent behavioral risk factors.

**Requires:** Run `cse160_project.Rmd` first in the same session so that
`wide`, `train`, `test`, `train_rf`, `test_rf`, `feature_cols`, `mlr_model`,
`mlr_rmse`, `mlr_r2`, `ridge_rmse`, `ridge_r2`, `rf_accuracy`, `y_train`,
`y_test`, `X_train`, `X_test`, and `cv_ridge` are all available.

```{r libraries, message=FALSE}
library(ggplot2)
library(glmnet)
library(randomForest)
```

## Reduced Feature Set

```{r no-cvd08-setup}
feature_cols_no_cvd08 <- feature_cols[feature_cols != "CVD08"]
cat("Dropped: CVD08\n")
cat("Remaining features:", length(feature_cols_no_cvd08), "\n")
print(feature_cols_no_cvd08)
```

## MLR Without CVD08

```{r no-cvd08-mlr}
mlr_formula_nc <- as.formula(paste("CVD09 ~", paste(feature_cols_no_cvd08, collapse = " + ")))
mlr_model_nc   <- lm(mlr_formula_nc, data = train)
mlr_pred_nc    <- predict(mlr_model_nc, newdata = test)

mlr_rmse_nc <- sqrt(mean((mlr_pred_nc - test$CVD09)^2))
mlr_r2_nc   <- 1 - sum((mlr_pred_nc - test$CVD09)^2) / sum((test$CVD09 - mean(test$CVD09))^2)

cat("MLR with    CVD08 — RMSE:", round(mlr_rmse,    3), "| R2:", round(mlr_r2,    3), "\n")
cat("MLR without CVD08 — RMSE:", round(mlr_rmse_nc, 3), "| R2:", round(mlr_r2_nc, 3), "\n")
```

## Ridge Without CVD08

```{r no-cvd08-ridge}
X_train_nc <- as.matrix(train[, feature_cols_no_cvd08])
X_test_nc  <- as.matrix(test[, feature_cols_no_cvd08])

set.seed(160)
cv_ridge_nc   <- cv.glmnet(X_train_nc, y_train, alpha = 0, nfolds = 10)
ridge_pred_nc <- predict(cv_ridge_nc, newx = X_test_nc, s = "lambda.min")

ridge_rmse_nc <- sqrt(mean((ridge_pred_nc - y_test)^2))
ridge_r2_nc   <- 1 - sum((ridge_pred_nc - y_test)^2) / sum((y_test - mean(y_test))^2)

cat("Ridge with    CVD08 — RMSE:", round(ridge_rmse,    3), "| R2:", round(ridge_r2,    3), "\n")
cat("Ridge without CVD08 — RMSE:", round(ridge_rmse_nc, 3), "| R2:", round(ridge_r2_nc, 3), "\n")
```

## Random Forest Without CVD08

```{r no-cvd08-rf}
set.seed(160)
rf_formula_nc  <- as.formula(paste("CVD09_range ~", paste(feature_cols_no_cvd08, collapse = " + ")))
rf_model_nc    <- randomForest(rf_formula_nc, data = train_rf, ntree = 500)
rf_pred_nc     <- predict(rf_model_nc, newdata = test_rf)
rf_accuracy_nc <- mean(rf_pred_nc == test_rf$CVD09_range)

cat("RF with    CVD08 — Accuracy:", round(rf_accuracy,    3), "\n")
cat("RF without CVD08 — Accuracy:", round(rf_accuracy_nc, 3), "\n")
```

## Comparison

```{r no-cvd08-comparison}
reg_comparison <- data.frame(
  Model = c("MLR", "MLR (no CVD08)", "Ridge", "Ridge (no CVD08)"),
  RMSE  = c(mlr_rmse,   mlr_rmse_nc,   ridge_rmse,   ridge_rmse_nc),
  R2    = c(mlr_r2,     mlr_r2_nc,     ridge_r2,     ridge_r2_nc)
)
cat("Regression models:\n")
print(reg_comparison)

cat("\nRandom Forest accuracy:\n")
cat("  With CVD08:   ", round(rf_accuracy,    3), "\n")
cat("  Without CVD08:", round(rf_accuracy_nc, 3), "\n")

ggplot(reg_comparison, aes(x = Model, y = RMSE, fill = Model)) +
  geom_col(show.legend = FALSE) +
  labs(
    title = "RMSE — With vs Without CVD08",
    x     = NULL,
    y     = "RMSE (deaths per 100k)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

ggplot(reg_comparison, aes(x = Model, y = R2, fill = Model)) +
  geom_col(show.legend = FALSE) +
  labs(
    title = "R-Squared — With vs Without CVD08",
    x     = NULL,
    y     = "R-Squared"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

rf_acc_df <- data.frame(
  Model    = c("RF (with CVD08)", "RF (no CVD08)"),
  Accuracy = c(rf_accuracy, rf_accuracy_nc)
)
ggplot(rf_acc_df, aes(x = Model, y = Accuracy, fill = Model)) +
  geom_col(show.legend = FALSE) +
  ylim(0, 1) +
  labs(
    title = "Random Forest Accuracy — With vs Without CVD08",
    x     = NULL,
    y     = "Accuracy"
  ) +
  theme_minimal()
```
